---
title: "Analyzing WeGo Data with R"
output: html_notebook
---

```{r}
library(tidyverse)
library(sf)
library(lubridate)
```

```{r}
headway <- read_csv("../data/headway_data_clean.csv")
```

**Part 1: Modeling Lateness**
Recall that a bus would be considered late if it is more than 6 minutes late (based on the adherence value). A bus is early if it is more than one minute early based on adherence.

*Negative (late) indicates departure time after scheduled time, and positive (early) indicates departure time before scheduled time.*

Start by doing some simple overall aggregates. 
*What percentage of the total stops are on time? How does this differ based on route? day of week? time of day?*
```{r}
headway |> 
  drop_na(ADHERENCE) |>
  summarise(ON_TIME = mean(ADHERENCE <= 1 & ADHERENCE >= -6)* 100, 
            LATE = mean(ADHERENCE <= -6)* 100, 
            EARLY = mean(ADHERENCE >= 1)* 100)
```
Create a new column based on whether a ride was on time or not. 
```{r}
headway <- headway |> 
  mutate(ON_TIME = ADHERENCE <= 1 & ADHERENCE >= -6)
```

```{r}
headway |> 
  drop_na(ADHERENCE) |> 
  summarise(ON_TIME = sum(ADHERENCE <= 1 & ADHERENCE >= -6, na.rm = TRUE)/ n(), 
            LATE = sum(ADHERENCE <= -6, na.rm = TRUE)/ n(), 
            EARLY = sum(ADHERENCE >= 1, na.rm = TRUE)/ n()) |> 
  mutate(
    ON_TIME = ON_TIME *100, 
    LATE = LATE * 100, 
    EARLY = EARLY * 100
  )
```

*How does this differ based on route?*
```{r}
headway |> 
  group_by(ROUTE_ABBR) |> 
  drop_na(ADHERENCE) |>
  summarise(ON_TIME = mean(ADHERENCE <= 1 & ADHERENCE >= -6)* 100, 
            LATE = mean(ADHERENCE <= -6)* 100, 
            EARLY = mean(ADHERENCE >= 1)* 100)
```

*How does this differ based on day of week?*
Create a new column that states the day of the week the ride occurred.
```{r}
headway <- headway |> 
  mutate(DAY_OF_WEEK = weekdays(DATE))
```

```{r}
headway |> 
  group_by(DAY_OF_WEEK) |> 
  drop_na(ADHERENCE) |>
  summarise(ON_TIME = mean(ADHERENCE <= 1 & ADHERENCE >= -6)* 100, 
            LATE = mean(ADHERENCE <= -6)* 100, 
            EARLY = mean(ADHERENCE >= 1)* 100)
```

*How does this differ based on time of day?*
Create a new column that states the day of the week the ride occurred.
```{r}
headway <- headway |> 
  mutate(HOUR_OF_DAY = hour(SCHEDULED_TIME))
```

```{r}
headway |> 
  group_by(HOUR_OF_DAY) |> 
  drop_na(ADHERENCE) |>
  summarise(ON_TIME = mean(ADHERENCE <= 1 & ADHERENCE >= -6)* 100, 
            LATE = mean(ADHERENCE <= -6)* 100, 
            EARLY = mean(ADHERENCE >= 1)* 100)
```

Now, consider the operators. 
*Which operators have the highest and lowest on-time percentages?*
```{r}
headway |> 
  group_by(OPERATOR) |> 
  drop_na(ADHERENCE) |>
  summarise(ON_TIME = mean(ADHERENCE <= 1 & ADHERENCE >= -6)* 100, 
            LATE = mean(ADHERENCE <= -6)* 100, 
            EARLY = mean(ADHERENCE >= 1)* 100) |> 
  filter(ON_TIME == min(ON_TIME) | ON_TIME == max(ON_TIME))
```

It's not really fair to compare operators with different number of observations with each other without taking into account the uncertainty due to small sample sizes. We could instead build a confidence interval around out estimates. Use the prop.test function to generate a confidence interval for the on-time percentage for each operator. Once you've done this, if we want to identify the best and worst operators in terms of on-time percentage you could look at those that have the highest lower bound and lowest upper bound, respectively.

```{r}

```


Now, let's take a modeling approach. Fit a multilevel logistic regression model with random effects for driver, where the target is whether a stop is on-time. Compare the operator-level random effects to what you computed using the confidence interval approach. WARNING: Make sure that OPERATOR is being treated as a factor.

We still haven't considered the fact that the route is likely going to influence the likelihood of being late. Fit a logistic regression model with a random term corresponding to the OPERATOR which account for the impact of route (and if you'd like, any other variables you want to control for) by including route as a fixed effect. WARNING: Make sure that you have converted OPERATOR and ROUTE_ABBR to factors before fitting this model.

Up to this point, we haven't considered the fact that observations within the same trip are likely going to be correlated. That is, if a driver is late at the first stop of a trip, they are probably going to be more likely to be late for other stops on the same route. As a result, our effective sample sizes are actually smaller for each operator. Determine for each operator the percentage of trips for which they are on-time, where on-time means that they were on time for all stops on that trip (or however you want to define an on-time trip). Does this change your evaluation of who the best and worst drivers are?

Try fitting a logistic regression model where for predicting whether a trip will be on time.

Another approach would be to go back to the full dataset (not aggregated by trip) and fit a model with trips nested within operators.
